Yeast RNA-seq 48 sample comparison

### 2018-01-10 datasets/RNA-Seq/yeast48-benchmarkFDR.Rmd
Add non-null comparisons

---
title: "Yeast RNA-seq 48 sample comparison"
author: "Keegan Korthauer"
date: "1/9/2018"
output: html_document
---

@@

### Introduction

@@

### Set up workspace

```{r, results='hide'}
# Load packages and source benchmark FDR
library(data.table)
library(tidyr)
library(dplyr)
library(readr)
library(ggplot2)
library(magrittr)
library(R.utils)
library(cowplot)
library(purrr)
library(DESeq2)

@@

### Data download

@@

```{r, readin}
files <- dir(path = "./data/", pattern = "*.bam.gbgout", full.names = TRUE)
sample_names <- sapply(strsplit(dir(path = "./data/", pattern = "*.bam.gbgout"), "_MID"),
                       function(x) x[[1]])

@@

### Analysis with DESeq2

@@

@@

@@

There are many thousand genes detected using the full set.

### Null Comparison

```{r, null}
# select a random subset of 20 WT samples
set.seed(85873)
dds_null <- dds[,colData(dds)$condition == "WT"]
dds_null <- dds_null[,sample(1:ncol(dds_null), 20)]

# add a fake condition column to coldat
colData(dds_null)$fake <-  factor(c(rep("WT", 10), rep("TRT", 10))[sample(1:20, 20)])
design(dds_null) <- ~fake

# results on full set
dds_null <- DESeq(dds_null)
resultsNames(dds_null) # lists the coefficients
resNULL <- results(dds_null, name="fake_WT_vs_TRT", independentFiltering = F)

### FDR Benchmark

@@

```{r}
strat_hist(geneExp, pvalue="pval", covariate="ind_covariate", maxy=2)
```

@@

>>>>

---
title: "Yeast RNA-seq 48 sample comparison"
author: "Keegan Korthauer"
output: 
    html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
---

@@

## Introduction

@@

## Set up workspace

```{r, results='hide', message=FALSE}
# Load packages and source benchmark FDR
library(data.table)
library(tidyr)
library(dplyr)
library(readr)
library(ggplot2)
library(magrittr)
library(R.utils)
library(cowplot)
library(purrr)
library(DESeq2)
library(truncnorm)

@@

## Data download

@@

```{r, readin, results='hide', message=FALSE}
files <- dir(path = "./yeast48-data/", pattern = "*.bam.gbgout", full.names = TRUE)
sample_names <- sapply(strsplit(dir(path = "./yeast48-data/", pattern = "*.bam.gbgout"), "_MID"),
                       function(x) x[[1]])

@@

## Analysis with DESeq2 & FDR Benchmarking

@@

### Full analysis (48 vs 48 non-null)

@@

# filter low count genes
counts <- counts[rowMeans(counts) > 1,]

@@

There are many thousand genes detected using the full set. Next, we'll build input data frame for summarized benchmark.

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resFULL), pval=resFULL$pvalue, SE=resFULL$lfcSE, 
                      ind_covariate = resFULL$baseMean, effect_size=resFULL$log2FoldChange, 
                      test_statistic=resFULL$stat))

@@

```{r, width=15, height=15}
strat_hist(geneExp, pvalue="pval", covariate="ind_covariate", maxy=32)
```

@@

```{r}
plotFDRMethodsOverlap(sb, alpha=0.05, nsets=ncol(sb), order.by="freq", decreasing=TRUE, supplementary=FALSE)

To do: compare this to results using a subset of samples from each condition.

### Null Comparison (10 versus 10)

Here we'll repeat the above, but for a null comparison 10 versus 10 Snf2 samples,
where the groups are selected randomly.

```{r}
# select a random subset of 20 WT samples
set.seed(8373)
dds_null <- dds[,colData(dds)$condition == "Snf2"]
dds_null <- dds_null[,sample(1:ncol(dds_null), 20)]

# add a fake condition column to coldat
colData(dds_null)$fake <-  factor(c(rep("A", 10), rep("B", 10))[sample(1:20, 20)])
design(dds_null) <- ~fake

# results 
dds_null <- DESeq(dds_null)
resultsNames(dds_null) # lists the coefficients
resNULL <- results(dds_null, name="fake_B_vs_A", independentFiltering = F)

geneExp <- geneExp %>% na.omit() %>% dplyr::filter(ind_covariate>=0)

```{r, width=15, height=15}
strat_hist(geneExp, pvalue="pval", covariate="ind_covariate", maxy=2)
```

### Non-null Comparison (10 versus 10 null with spike-in)

```{r}
nDE <- 500

# pick random set of nDE genes to add signal to
DE <- sample(1:nrow(dds_null), nDE)
truth <- rep(FALSE, nrow(dds_null))
truth[DE] <- TRUE
names(truth) <- rownames(dds_null)

# randomly sample a log2FC from truncated N(0,0.5) (greater than 0.5) 
log2FC <- rep(0, nrow(dds_null))
log2FC[DE] <- rtruncnorm(nDE, 0.5, sd=0.5) * sample(c(1,-1), size=nDE, replace=TRUE)
counts_new <- counts(dds_null)
counts_new[,colData(dds_null)$fake=="A"] <- counts(dds_null)[,colData(dds_null)$fake=="A"] * 2^log2FC
counts_new <- apply(counts_new, 2, as.integer)
counts(dds_null) <- counts_new

# results 
dds_null <- DESeq(dds_null)
resultsNames(dds_null) # lists the coefficients
resNULL <- results(dds_null, name="fake_B_vs_A", independentFiltering = F)

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resNULL), pval=resNULL$pvalue, SE=resNULL$lfcSE, 
                      ind_covariate = resNULL$baseMean, effect_size=resNULL$log2FoldChange, 
                      test_statistic=resNULL$stat,
                      qvalue=truth))

geneExp <- geneExp %>% na.omit() %>% dplyr::filter(ind_covariate>=0)

```{r, width=15, height=15}
strat_hist(geneExp, pvalue="pval", covariate="ind_covariate", maxy=4)
```

Run benchmark methods and examine ROC curve.

```{r}
bd <- initializeBenchDesign()
sb <- bd %>% buildBench(data=geneExp, parallel = T, truthCols = "qvalue")

sb <- addDefaultMetrics(sb)
pf <- estimatePerformanceMetrics(sb, alpha = c(0.001, 0.01, 0.025, 0.05, 0.1, 0.2), tidy=TRUE) 
tpr <- pf %>% filter(performanceMetric == "TPR") %>% 
  mutate(TPR=value) %>% 
  select(blabel, alpha, TPR)
fpr <- pf %>% filter(performanceMetric == "FPR") %>% 
  mutate(FPR=value) %>% 
  select(blabel, alpha, FPR)
roc <- left_join(tpr, fpr) %>%
  filter(blabel %in% c("ashs", "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf",
             "scott-empirical", "scott-theoretical"))

ggplot(data=roc,aes(FPR, TPR, color=blabel)) +
  geom_point() +
  geom_line() +
  xlim(0,0.2)
```

```{r}
plotFDRMethodsOverlap(sb, alpha=0.05, nsets=ncol(sb), order.by="freq", decreasing=TRUE, supplementary=FALSE)

### Null Comparison (5 versus 5)

Again we'll look at a null comparison of wild type samples in artifical groups, but
for a smaller sample size (5 versus 5).

```{r}
# select a random subset of 10 WT samples
dds_null <- dds[,colData(dds)$condition == "Snf2"]
dds_null <- dds_null[,sample(1:ncol(dds_null), 10)]

# add a fake condition column to coldat
colData(dds_null)$fake <-  factor(c(rep("A", 5), rep("B", 5))[sample(1:10, 10)])
design(dds_null) <- ~fake

# results
dds_null <- DESeq(dds_null)
resultsNames(dds_null) # lists the coefficients
resNULL <- results(dds_null, name="fake_B_vs_A", independentFiltering = F)

```{r, width=15, height=15}
strat_hist(geneExp, pvalue="pval", covariate="ind_covariate", maxy=3)
```

```{r}
#plotFDRMethodsOverlap(sb, alpha=0.05, nsets=ncol(sb), order.by="freq", decreasing=TRUE, supplementary=FALSE)
```

### Non-null Comparison (5 versus 5 null with spike-in)

```{r}
nDE <- 500

# pick random set of nDE genes to add signal to
DE <- sample(1:nrow(dds_null), nDE)
truth <- rep(FALSE, nrow(dds_null))
truth[DE] <- TRUE
names(truth) <- rownames(dds_null)

# randomly sample a log2FC from truncated N(0,0.5) (greater than 0.5) 
log2FC <- rep(0, nrow(dds_null))
log2FC[DE] <- rtruncnorm(nDE, 0.5, sd=0.5) * sample(c(1,-1), size=nDE, replace=TRUE)
counts_new <- counts(dds_null)
counts_new[,colData(dds_null)$fake=="A"] <- counts(dds_null)[,colData(dds_null)$fake=="A"] * 2^log2FC
counts_new <- apply(counts_new, 2, as.integer)
counts(dds_null) <- counts_new

# results 
dds_null <- DESeq(dds_null)
resultsNames(dds_null) # lists the coefficients
resNULL <- results(dds_null, name="fake_B_vs_A", independentFiltering = F)

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resNULL), pval=resNULL$pvalue, SE=resNULL$lfcSE, 
                      ind_covariate = resNULL$baseMean, effect_size=resNULL$log2FoldChange, 
                      test_statistic=resNULL$stat,
                      qvalue=truth))

```{r,width=15, height=15}
strat_hist(geneExp, pvalue="pval", covariate="ind_covariate", maxy=4)
```

Run benchmark methods and examine ROC curve.

```{r}
bd <- initializeBenchDesign()
sb <- bd %>% buildBench(data=geneExp, parallel = T, truthCols = "qvalue")

sb <- addDefaultMetrics(sb)
pf <- estimatePerformanceMetrics(sb, alpha = c(0.001, 0.01, 0.025, 0.05, 0.1, 0.2), tidy=TRUE) 
tpr <- pf %>% filter(performanceMetric == "TPR") %>% 
  mutate(TPR=value) %>% 
  select(blabel, alpha, TPR)
fpr <- pf %>% filter(performanceMetric == "FPR") %>% 
  mutate(FPR=value) %>% 
  select(blabel, alpha, FPR)
roc <- left_join(tpr, fpr) %>%
  filter(blabel %in% c("ashs", "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf",
             "scott-empirical", "scott-theoretical"))

ggplot(data=roc,aes(FPR, TPR, color=blabel)) +
  geom_point() +
  geom_line() + 
  xlim(0,0.2)
```
